---
title: "Bike_share_case_study_md"
author: "Jordan Creenaune"
date: "8/21/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Google Data Analytics Certificate - Track 1 - Case study one.

This is a guided case study from module 8 of the Google Data Analytics Certificate. 

Essential question - Case Study: How Does a Bike-Share Navigate Speedy Success? 

  - How are the different types of memberships using bikeshare programs represented in the data? 
  - What information can we draw from historical data that can help us understand the difference between casual riders and riders      with memberships?

##STEP 1 - Install required packages 

```{r}
library(tidyverse)  #A series of packages used to wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)    #We will perform visualisations using this package
```

##STEP 2 - Import data sets and combine
Set working directory and import datasets 
```{r}
setwd("/Users/jordancreenaune/Documents/R")

df_202007 <- read_csv("202007-divvy-tripdata.csv")
df_202008 <- read_csv("202008-divvy-tripdata.csv")
df_202009 <- read_csv("202009-divvy-tripdata.csv")
df_202010 <- read_csv("202010-divvy-tripdata.csv")
df_202011 <- read_csv("202011-divvy-tripdata.csv")
df_202012 <- read_csv("202012-divvy-tripdata.csv")
df_202101 <- read_csv("202101-divvy-tripdata.csv")
df_202102 <- read_csv("202102-divvy-tripdata.csv")
df_202103 <- read_csv("202103-divvy-tripdata.csv")
df_202104 <- read_csv("202104-divvy-tripdata.csv")
df_202105 <- read_csv("202105-divvy-tripdata.csv")
df_202106 <- read_csv("202106-divvy-tripdata.csv")

```


Combine datasets into a single df
```{r}
tripdata <- rbind(df_202007,
                  df_202008,
                  df_202009,
                  df_202010,
                  df_202011,
                  df_202012,
                  df_202101,
                  df_202102,
                  df_202103,
                  df_202104,
                  df_202105,
                  df_202106)
glimpse(tripdata)
```

##STEP 3- Clean Datasets
Clean data - remove values that contain missing components in the data 

```{r}

# Remove rows with missing values
colSums(is.na(tripdata))

# 5% of data with missing values will be removed
tripdata_cleaned <- tripdata[complete.cases(tripdata), ]

# data with started_at greater than ended_at will be removed - remove possible inconsistencies
tripdata_cleaned <- tripdata_cleaned %>% 
  filter(tripdata_cleaned$started_at < tripdata_cleaned$ended_at)

```

##STEP 4 - Analyse and manipulate data 
Create new column `ride_length` - Find the difference between ended_at and started_at 

```{r}
tripdata_cleaned$ride_length <- (tripdata_cleaned$ended_at - tripdata_cleaned$started_at)
head(tripdata_cleaned)
```

Plot - demonstrating casual vs member riders for this data set 

```{r}

number_of_rides <- tripdata_cleaned %>% 
                      group_by(member_casual) %>% 
                      summarize(number_of_rides=n())


ggplot(number_of_rides, aes(x=member_casual, y=number_of_rides)) + 
  geom_bar(stat = "identity",fill="dodgerblue3") + 
  labs(title = "Number of Rides July 2020 - July 2021") +
  ylab("Number of rides")
  
```


Create 2 new columns `day_of_week` and 'day' using the lubridate package

```{r}
tripdata_cleaned$day_of_week <- wday(tripdata_cleaned$started_at, label = FALSE)
tripdata_cleaned$day <- weekdays(as.Date(tripdata_cleaned$started_at))

 
```

Find mean - max - min - mode of ride_length 

```{r}
# mean of ride_length
mean_ride_length <- tripdata_cleaned %>% 
  summarize(mean(ride_length))

# max ride_length
max_ride_length <- tripdata_cleaned %>% 
  summarize(max(ride_length))

# min ride_length
min_ride_length <- tripdata_cleaned %>% 
  summarize(min(ride_length))

#Mode of Column 
#Create a mode function 
getmode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

# Calculate the mode of day_of_week using the Mode function (from above)
day_mode <- getmode(tripdata_cleaned$day)
print(day_mode)
day_of_week_mode <- getmode(tripdata_cleaned$day_of_week)
#print(day_of_week_mode)

data_summary <- data.frame(max_ride_length,min_ride_length,mean_ride_length,day_of_week_mode)
print(data_summary)

```

Average ride length between member and casual riders in seconds 
```{r}
# average ride_length for members and casual riders
tripdata_cleaned %>% 
  group_by(member_casual) %>% 
  summarize(mean(ride_length))
```

Average ride_length for users by day_of_week - inclusive of both membership types

```{r}
average_d_o_w <- tripdata_cleaned %>% 
                    group_by(day_of_week,day) %>% 
                    summarize(mean(ride_length))

average_d_o_w$mean_ride_length <- round(average_d_o_w$`mean(ride_length)` ,digit=2)
average_d_o_w <- average_d_o_w[ -c(3) ]

average_d_o_w
```

##STEP 5- Visualisation 

Visualisation of time series data
Order data by date and time, isolate one column

```{r}
date_order <- tripdata_cleaned[order(tripdata_cleaned$started_at),3,drop=FALSE ]
head(date_order) #Check data is in the correct order 
```


Smooth line plot - demonstrates the amount of rides over the course of the year. 

```{r}
date_order_plot <-  date_order %>%  
  group_by(started_at) %>% 
  summarise(session_count = n()) %>% 
  ggplot(aes(started_at, session_count)) + 
  geom_smooth()
  labs(title = "Number of Rides July 2020 - July 2021",
      subtitle = "Hundreds of thousands of rides") +
  ylab("Number_of_Rides - Hundreds of thousands") +
  xlab("time_of_year")

date_order_plot

```

Using this visualisation - we can clearly see that through both memberships and casual users, there are clear trends throughout the year for the use of the bikeshare program. As a result of the geographical location, Chicago, there is a significant decline in the winter months and many more users in the summer months. 

Number of rides for users by day_of_week

```{r}
tripdata_cleaned %>% 
  group_by(ride_id, day_of_week) %>% 
  summarize(number_of_rides=n())

head(tripdata_cleaned)
# average ride_length by type and day of week
counts <- aggregate(tripdata_cleaned$ride_length ~ tripdata_cleaned$member_casual +
                      tripdata_cleaned$day_of_week + tripdata_cleaned$day, FUN = mean)
                      names(counts)[1] <- "member_casual"
                      names(counts)[2] <- "day_of_week"
                      names(counts)[3] <- "day"
                      names(counts)[4] <- "mean_ride_length"
                      
                      counts$mean_ride_length <- round(counts$mean_ride_length,digit=2)
counts <- counts[ -c(2) ]
counts$day <- factor(counts$day, levels= c("Sunday", "Monday", 
                                         "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
counts <- counts[order(counts$day), ]

head(counts)

```

The above data frame demonstrates the mean_ride_length with regard to day of the week and the membership type (casual or member). We can see from the head of this dataframe that casual riders have significantly longer rides than members that encompasses any day of the week.



Average ride time by each day for members vs casual users

```{r}
average_casual_vs_member <- 
  aggregate(tripdata_cleaned$ride_length ~ tripdata_cleaned$member_casual + tripdata_cleaned$day, FUN = mean)
  names(average_casual_vs_member)[1] <- "member_casual"
  names(average_casual_vs_member)[2] <- "day_of_week"
  names(average_casual_vs_member)[3] <- "mean_ride_length"
  average_casual_vs_member$mean_ride_length <- round(average_casual_vs_member$`mean_ride_length` ,digit=2)
printrows <- average_casual_vs_member[1:14,]
printrows

```

The above dataframe demonstrates the mean ride_length between casual and member riders. It is clear through this dataframe that casual riders have significantly longer duration of rides than riders that hold memberships. 


Analyze and visualise ridership data by type and weekday

```{r}
tripdata_cleaned %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarize(number_of_rides = n(),
            average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)


# visualize number of rides by rider type
tripdata_cleaned %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarize(number_of_rides = n(),
            average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#CC6633","#6699CC")) +
  labs(title = "Number of Rides by Days and Rider Type",
       subtitle = "Members versus Casual Users") +
  ylab("Number of  - Hundreds of thousands") +
  xlab("Day of Week")

```


This visusalistion demonstrates the number of rides between casual and those who hold memberships. It is clear that during the week, riders who hold memberships dominate usage throughout the work week (Monday to Friday). Those riders who are casual riders have more riders throughout weekends and what we have previously learned that they also have a longer ride duration. 



Visualization for average duration - with regard to membership status (casual vs member)
```{r}
tripdata_cleaned %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarize(average_duration = mean(ride_length)/60) %>% 
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#CC6633","#6699CC")) +
  labs(title = "Average Duration of Rides by Days and Rider Type",
       subtitle = "Members versus Casual Users in minutes") +
  ylab("Average Duration of Rides- minutes") +
  xlab("Day of Week")
```

This visualisation demonstrates the average duration of rides with regard to days and rider type. It is evident through this analysis that casual riders have significantly longer rides than members each day. Combined with information that we have already learned about membership types, members are more likely to ride more often during the week and have shorter rides. This could be due to the convenience and the ease at which they are able to find and rent a bike to get to their destination. 


Average ride_length and type and month

```{r}
tripdata_cleaned$month <- month(tripdata_cleaned$started_at, label = TRUE)
rides <- aggregate(tripdata_cleaned$ride_length ~ tripdata_cleaned$member_casual +
                     tripdata_cleaned$month,FUN = mean)
```

The ways in which casual and member riders use this service are different in a variety of ways. The last two graphs will demonstrate how these types of riders behave in terms of the time of day that they're using the service. 

Casual Riders time of day 
```{r}
casual_rider <- subset(tripdata_cleaned, member_casual == "casual")
#separate date and time into two columns 
casual_time <- separate(casual_rider, started_at, into = c("date", "time"), sep = " ",)
#Drop non essential columns 
casual_rider_time <- casual_time[order(casual_time$time),4,drop=FALSE ]
#Aggregate time and count instances in a new column
casual_time_count <- casual_rider_time %>% count(time)



member_rider <- subset(tripdata_cleaned, member_casual == "member")
#separate date and time into two columns 
member_time <- separate(member_rider, started_at, into = c("date", "time"), sep = " ",)
#Drop non essential columns 
member_rider_time <- member_time[order(member_time$time),4,drop=FALSE ]
#Aggregate time and count instances in a new column
member_time_count <- member_rider_time %>% count(time)

ggplot() +
  geom_point(data=casual_time_count, aes(time, n,, color= 'casual_rider')) +
  geom_point(data=member_time_count, aes(time, n, color= 'member_rider'))

```





Dataset is ordered and exported to a csv for further analysis to be imported to tableau or PowerBI

```{r}
alltrips <- tripdata_cleaned %>% 
  select(-day_of_week)
alltrips$day_of_week <- wday(alltrips$started_at, label = TRUE)

alltrips_ordered <- alltrips[order(alltrips$started_at),]

head(alltrips_ordered )
#write.csv(alltrips_ordered, file = "all_trips.csv", row.names = FALSE)

```



